Rheology of Cubic Blue Phases

### 2013-06-28 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2029 40876441-3496-0410-a613-a2fca35b2ae2

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up \rev{into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction according to the imposed flow}.
BPI leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. \rev{Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states. For even
larger flow rates a Grandjean texture is observed, just as in the case of BPII.
At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.}
Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the the \dag symbol from the title and the above footnotetext.

@@

%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
\rev{Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.}
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. \rev{These result, however, were derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma} l$. \rev{The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.}

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
\rev{For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory response (PRC), 
amorphous networks (AN), amorphous and metastable states (AMS) as well as those featuring a Grandjean texture (GJ) and 
a flow-aligned nematic state (FAN). The last two occur regardless of the initial state. 
They are for Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) and
for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
BPI-3 (AMS) ($3.125\e{-4}\lesssim \gd\lesssim 7.813\e{-4}; 2.67 \lesssim {\it Er} \lesssim 6.01$),
BPI-4 (GJ) ($9.375\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 8\lesssim {\it Er} \lesssim 16)$ and
BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.34\lesssim {\it Er}$).}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
\caption{\rev{Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.}}
\label{bp2-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{\rev{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ (normalised to the background viscosity $\eta$) 
of BPII versus strain $\gamma = \gd\cdot t$ in the regimes BPII-1 (top, PRC) and BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.}} 
\label{bp2-appvisc}
\end{figure}

@@

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
remarkably small in BPII, in the range of shear rates which we have explored here.}

\rev{Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence at very low shear rate of a permeation mode is probably due to our
choice of boundary conditions.}

@@

\rev{We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.}

\rev{The picture at the bottom of Fig. \ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
the flow-aligned nematic state. We will address these regimes in section \ref{cholflow}, as they are common to BPI and BPII.}

\rev{The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
to the tumbling motion of the local director field, which is absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.}

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. \rev{Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
whereas the network as a whole moves only in one direction.}

@@

\rev{Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve}. 
The plot shows that there is more shear thinning in BPI than in BPII, for
the shear rate range we considered.}

For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
response versus time, and in the network morphology -- we label this 
regime BPI-2. 
\rev{This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{For these very low shear rates the flow curve in Fig. \ref{bp1-flowcurve}
develops a horizontal branch, which suggest the existence of a yield stress.} 
We refer to this flow regime as BPI-1.

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value 
\rev{in the region of Ericksen numbers $Er \simeq O(1)$
where viscous and elastic forces are of the same magnitude. 
Close to this breakup point the network seems to become
increasingly sensitive towards small perturbations in the flow field
that cause irregularities in the order parameter structure.
These distortions couple back to the flow and form a possible
route towards "rheochaos"~\cite{rheochaos,Cates:2002}.}

We also observe that

regime BPI-3 ($3.125\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $2.67 \lesssim {\it Er} \lesssim 6.01$)

Finally, for shear rates such that
$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
configuration with the helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{cholflow}).

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
\rev{Depending on the configuration in steady shear flow five different regimes 
can be identified: amorphous BP network with yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
breakup of the disclination network into amorphous state (BPI-3, open circles)
and metastable configurations (BPI-3, solid circles), 
cholesteric helix with axis in gradient direction (BPI-4, open triangles) 
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.}}
\label{bp1-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
and strain (bottom).}
\label{bp1-appvisc}
\end{figure}

@@

To gain further insight into the flow-induced deformation, and eventual
breakdown, of BPI, it is instructive to follow the dynamics of 
the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
Three different stages can be distinguished. 
Just after the onset of the shear flow, the disclination lines 
in BPI get more and more squeezed together, and this is incompatible with
the defect topology in this phase.
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the shear transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-1-disc} (top row).
At this point the original BPI has already been significantly deformed.
Shortly afterwards regular oscillations set up temporarily,
where the disclinations form double helices which tilt and realign under the shear.
There is no perceivable movement of the 
network along the vorticity direction at any stage of the cycle.
(This contrasts with the BPII-2 results reported earlier.)

>>>>

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction according to the imposed flow.
}
BPI leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states. For even
larger flow rates a Grandjean texture is observed, just as in the case of BPII.
At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the \dag symbol from the title and the above footnotetext.

@@

%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the correct place in the list.}

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
\rev{
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
}
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
\rev{
These result, however, were derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma} l$. 
\rev{
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
}

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
\rev{
For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
GJ and FAN occur regardless of the initial state.
}

\rev{
These regimes are characterised by the following flow rates: 
For Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) 
and for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
BPI-3 (AN) ($3.125\e{-4}\lesssim \gd\lesssim 4.69\e{-4}; 2.67 \lesssim {\it Er} \lesssim 4$),
BPI-4 (GJ and FGJ) ($5.47\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 4.67\lesssim {\it Er} \lesssim 16)$ and
BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.33\lesssim {\it Er}$).
}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
\caption{
\rev{
Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
}
}
\label{bp2-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
\rev{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
}
} 
\label{bp2-appvisc}
\end{figure}

@@

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure. The coefficients of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
remarkably small in BPII, in the range of shear rates which we have explored here.}

\rev{
Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence at very low shear rate of a permeation mode is probably due to our
choice of boundary conditions.
}

@@

\rev{
We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
}

\rev{
The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
}

\rev{
The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
to the tumbling motion of the local director field, which is absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
}

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
\rev{
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
whereas the network as a whole moves only in one direction.
}

@@

\rev{
Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve}, whereas Fig. \ref{bp1-appvisc} shows 
the apparent viscosity ration $\eta_{app}/\eta$ over total strain.
}

For intermediate (but not low) shear rates
($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more find regular oscillations in the stress
response versus time, and in the network morphology -- we label this 
regime BPI-2. 
\rev{
This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.
}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{
For these very low shear rates 
($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
the flow curve in Fig. \ref{bp1-flowcurve}
apparently develops a horizontal yield stress branch. 
We refer to this flow regime as BPI-1.
}

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value 
\rev{
in the region of Ericksen numbers $Er \simeq O(1)$.
This is a regime 
($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
where viscous and elastic forces have more or less 
the same magnitude we labelled BPI-3. 
In this window of strain rates the network seems to become
increasingly sensitive towards small perturbations in the flow field
that can cause irregularities in the order parameter structure.
Due to the back flow, i.e. the coupling of flow and order in our model, 
these distortions form a possible route towards an instance of 
"rheochaos"~\cite{rheochaos,Cates:2002}.
}

\rev{
Finally, for shear rates such that
$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
and before breakup into a flow-aligned nematic occurs 
(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
we observe the same cholesteric Grandjean configuration with the 
helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{gj-fan}). This regime is referred to as BPI-4.
}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
\rev{
Depending on the configuration in steady shear flow five different regimes 
can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
breakup of the disclination network into amorphous state (BPI-3, open circles)
\com{Calculate another shear rate to get a regime},
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture and frustrated metastable states that transform into 
a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{
\rev{
Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
The top picture shows regime BPI-1 (AN) where
the BP network transforms into an amorphous network featuring a yield stress.
The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
whereas the picture at the bottom
depicts data of the Grandjean texture (GJ and FGJ if frustrated) in regime BPI-4 
and the flow-aligned nematic state (BPI-5, FAN). 
Note that the $\eta_{app}/\eta$ is negative for the lowest 
flow rates as the (negative) contribution of the elastic forces
of the network dominates the total stress.
}
}
\label{bp1-appvisc}
\end{figure}

@@

To gain further insight into the flow-induced deformation, and eventual
breakdown, of BPI, it is instructive to follow the dynamics of 
the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
Three different stages can be distinguished. 
Just after the onset of the shear flow, the disclination lines 
in BPI get more and more squeezed together, and this is incompatible with
the defect topology in this phase.
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the shear transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-1-disc} (top row).
At this point the original BPI has already been significantly deformed.
Shortly afterwards regular oscillations set up temporarily,
where the disclinations form double helices which tilt and realign under the shear.

This mode of flow proves unstable, as after a few cycles 
distortions appear which lead to further destabilisation -- presumably in
view of the large stress fluctuations discussed above. 
Finally, the system loses order and 
transforms into an amorphous network state with almost constant
stress in the region of $\Pi_{xy}\simeq 1-2\e{-5}$ LBU.
If the shear is stopped, this flow-induced amorphous state
remains arrested and metastable, and cannot
find its way back to the original BPI structure.
